Trends from 30-year observations of downward solar irradiance in Thessaloniki, Greece

### 2023-12-02 MDPI_submission.Rmd
Commit 2023-12-02 22:14

```{r echo=F, include=T, warning=FALSE, message=FALSE}
suppressMessages(
    res <- rmarkdown::find_pandoc(dir = '/usr/lib/rstudio/resources/app/bin/quarto/bin/tools')
)
## document configurations
knitr::opts_chunk$set(comment    = ""      )
# knitr::opts_chunk$set(dev        = "pdf"   )
knitr::opts_chunk$set(out.width  = "70%"   )
knitr::opts_chunk$set(fig.align  = "center")
knitr::opts_chunk$set(fig.pos    = 'h!'    )
options(knitr.graphics.auto_pdf = FALSE)
library(pander)
library(data.table)
library(kableExtra)
library(dplyr)
## load data from trends analysis
source("./DHI_GHI_00_raw_data.R")
# load(common_data)
load("./data/Input_1_longterm_trends.Rda")
## number of measurements for each dataset
SDR_N_All   <- sum(  ALL_1_daily_DESEAS[,GLB_att_N])
SDR_N_Clear <- sum(CLEAR_1_daily_DESEAS[,GLB_att_N])
SDR_N_Cloud <- sum(CLOUD_1_daily_DESEAS[,GLB_att_N])
## TSI data span info
TSIinfo <- data.table(read.csv("./figures/tbl_tsi_info.dat",
                    strip.white = TRUE,
                    sep = ";",
                    header = TRUE))
## variables we use in the analysis
source("./DHI_GHI_0_variables.R")

>>>>

```{r echo=F, include=T, warning=FALSE, message=FALSE}
suppressMessages(
    res <- rmarkdown::find_pandoc(dir = '/usr/lib/rstudio/resources/app/bin/quarto/bin/tools')
)
## document configurations
knitr::opts_chunk$set(comment    = ""      )
# knitr::opts_chunk$set(dev        = "pdf"   )
knitr::opts_chunk$set(out.width  = "70%"   )
knitr::opts_chunk$set(fig.align  = "center")
knitr::opts_chunk$set(fig.pos    = 'h!'    )
options(knitr.graphics.auto_pdf = FALSE)
library(pander)
library(data.table)
library(kableExtra)
library(dplyr)
## load data from trends analysis
# source("./DHI_GHI_00_raw_data.R")
# load(common_data)
load("./data/Input_1_longterm_trends.Rda")
## number of measurements for each dataset
SDR_N_All   <- sum(  ALL_1_daily_DESEAS[,GLB_att_N])
SDR_N_Clear <- sum(CLEAR_1_daily_DESEAS[,GLB_att_N])
SDR_N_Cloud <- sum(CLOUD_1_daily_DESEAS[,GLB_att_N])
## TSI data span info
TSIinfo <- data.table(read.csv("./figures/tbl_tsi_info.dat",
                    strip.white = TRUE,
                    sep = ";",
                    header = TRUE))
## variables we use in the analysis
source("./DHI_GHI_0_variables.R")

### 2023-12-02 MDPI_submission.tex
Commit 2023-12-01 18:14
Commit 2023-12-02 22:14

Due to the significant measurement uncertainty when the Sun is near the
horizon, we have excluded all measurements with SZA greater than
\(85^\circ\). Moreover, due to some obstructions around the site (hills
and buildings), we excluded data with azimuth angle between \(58^\circ\)
and \(120^\circ\) with SZA greater than \(78^\circ\). On the latter
instances, Sun is systematically not visible from the instrument's
location. To make the measurements comparable throughout the dataset, we
adjusted all one-minute radiometric values to the mean Sun - Earth
distance. Subsequently, we made all measurements relative to the Total
Solar Irradiance (TSI) at \(1\,\text{au}\), in order to compensate for
the Sun's intensity variability, using a time series of satellite TSI
observations. The TSI data we used are part of the ``NOAA Climate Data
Record of Total Solar Irradiance'' dataset \citep{Coddington2005}. The
initial daily values of this dataset were interpolated to match the time
step of our measurements. In the final dataset \(40.1\%\) of the data
were identified as under clear-sky conditions and \(59.9\%\) as under
cloud-sky conditions.

@@

We calculated the linear SDR trends, from the departures of the mean
daily values from the daily climatology and for the three sky conditions
(Table~\ref{tab:trendtable}). In Figure~\ref{fig:trendALL} we present
only the time series under all-sky conditions; the plots for clear-sky
and cloud-sky conditions, are very similar and are shown in the Appendix
(Figures~\ref{fig:trendCLEAR} and~ \ref{fig:trendCLOUD}). We observe a
positive trend for all-sky conditions (\(0.38\,\%/y\)), a very close but
smaller trend for clear-skies (\(0.16\,\%/y\)) and a negative weaker
trend for cloudy-skies (\(0.019\,\%/y\)). In the studied period, there
is no significant break or change in the variability of the time series.
Other studies for the European region reported a change of the SDR
slope, around 1980 \citep{Wild2021, Yuan2021, Ohmura2009}, a few years
before the start of our records. It is interesting to note, that for the
observations period, the trend of the TSI is \(-0.00022\,\%/y\), and
thus we can not attribute any major effect on SDR trend to Solar
variability.

@@

\caption{\label{tab:trendtable}Trends in SDR daily means for different sky conditions for the period 1993 - 2023.}
\begin{tabu} to \linewidth {>{\centering\arraybackslash}p{8em}>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\centering}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\centering}X}
\toprule
slope & slope.p & Rsqrd & RsqrdAdj & cor.t & cor.esimate & DATA & N & slope.stat\_sig\\
\midrule
0.376 & 0.00e+00 & 0.0083302 & 0.0082334 & 9.279 & 0.091270 & All skies & 10252 & 100.00\\
0.156 & 7.00e-07 & 0.0030928 & 0.0029676 & 4.971 & 0.055613 & Clear skies & 7967 & 100.00\\
0.019 & 7.23e-01 & 0.0000123 & -0.0000858 & 0.354 & 0.003506 & Cloudy skies & 10194 & 27.66\\
\bottomrule
\end{tabu}
\end{table}

@@

Similarly to the long term trends discussed above, we have calculated
the trend of the anomalies for the three different sky conditions, and
for each season of the year, using the corresponding mean monthly values
(Figure~\ref{fig:seasonalALL} and Table~\ref{tab:trendseasontable}). For
all-sky conditions the trend in SDR in winter is the largest
(\(0.69\,\%/y\)), followed by the trend in autumn (\(0.43\,\%/y\), a
value close to the long term trend) both statistically significant above
the \(99\,\%\) confidence level. In spring and summer, the trends are
much smaller and of lesser statistical significance. These seasonal
differences indicate a possible relation of the trends in SDR to trends
of clouds during winter and autumn. For clear-skies, the trend in winter
is \(0.7\,\%/y\), larger than for all-skies (\(0.69\,\%/y\)), which is
another indication of a decreasing trend in cloud optical thickness.
Moreover, the trends under clear- and cloudy-sky conditions are almost
complementary to each other, particularly for winter and autumn, where
the signal is stronger. During spring and summer the statistical
significance is very low and the actual trend too small for a meaningful
comparison.

@@

\caption{\label{tab:trendseasontable}SDR linear trends of monthly anomalies for each season of the year.}
\begin{tabu} to \linewidth {>{\centering\arraybackslash}p{8em}>{\centering}X>{\raggedleft}X>{\raggedleft}X>{\centering}X>{\centering}X>{\raggedleft}X>{\raggedleft}X>{\centering}X}
\toprule
Trend [\%/year] & slope.p & Rsqrd & RsqrdAdj & cor.t & cor.esimate & Sky condition & Season & slope.stat\_sig\\
\midrule
\cellcolor{gray!6}{0.6870} & \cellcolor{gray!6}{1.33e-03} & \cellcolor{gray!6}{0.3121839} & \cellcolor{gray!6}{0.2876190} & \cellcolor{gray!6}{3.5649059} & \cellcolor{gray!6}{0.5587342} & \cellcolor{gray!6}{All skies} & \cellcolor{gray!6}{Winter} & \cellcolor{gray!6}{99.90}\\
\cmidrule{1-9}
0.1460 & 1.78e-01 & 0.0616942 & 0.0293388 & 1.3808581 & 0.2483832 & All skies & Spring & 82.20\\
\cmidrule{1-9}
\cellcolor{gray!6}{0.1230} & \cellcolor{gray!6}{9.61e-02} & \cellcolor{gray!6}{0.0957525} & \cellcolor{gray!6}{0.0634580} & \cellcolor{gray!6}{1.7219100} & \cellcolor{gray!6}{0.3094390} & \cellcolor{gray!6}{All skies} & \cellcolor{gray!6}{Summer} & \cellcolor{gray!6}{90.40}\\
\cmidrule{1-9}
0.4300 & 6.67e-03 & 0.2347186 & 0.2073871 & 2.9305032 & 0.4844777 & All skies & Autumn & 99.30\\
\cmidrule{1-9}
\cellcolor{gray!6}{0.6990} & \cellcolor{gray!6}{8.51e-05} & \cellcolor{gray!6}{0.4293113} & \cellcolor{gray!6}{0.4089296} & \cellcolor{gray!6}{4.5895019} & \cellcolor{gray!6}{0.6552185} & \cellcolor{gray!6}{Clear skies} & \cellcolor{gray!6}{Winter} & \cellcolor{gray!6}{100.00}\\
\cmidrule{1-9}
-0.0265 & 8.46e-01 & 0.0013302 & -0.0331066 & -0.1965409 & -0.0364724 & Clear skies & Spring & 15.40\\
\cmidrule{1-9}
\cellcolor{gray!6}{-0.1810} & \cellcolor{gray!6}{8.08e-03} & \cellcolor{gray!6}{0.2250803} & \cellcolor{gray!6}{0.1974046} & \cellcolor{gray!6}{-2.8518021} & \cellcolor{gray!6}{-0.4744263} & \cellcolor{gray!6}{Clear skies} & \cellcolor{gray!6}{Summer} & \cellcolor{gray!6}{99.20}\\
\cmidrule{1-9}
0.1260 & 2.57e-01 & 0.0455591 & 0.0114719 & 1.1560907 & 0.2134457 & Clear skies & Autumn & 74.30\\
\cmidrule{1-9}
\cellcolor{gray!6}{0.0944} & \cellcolor{gray!6}{6.13e-01} & \cellcolor{gray!6}{0.0092345} & \cellcolor{gray!6}{-0.0261500} & \cellcolor{gray!6}{0.5108573} & \cellcolor{gray!6}{0.0960962} & \cellcolor{gray!6}{Cloudy skies} & \cellcolor{gray!6}{Winter} & \cellcolor{gray!6}{38.70}\\
\cmidrule{1-9}
0.0161 & 9.09e-01 & 0.0004533 & -0.0340138 & 0.1146809 & 0.0212909 & Cloudy skies & Spring & 9.05\\
\cmidrule{1-9}
\cellcolor{gray!6}{-0.0283} & \cellcolor{gray!6}{8.07e-01} & \cellcolor{gray!6}{0.0021748} & \cellcolor{gray!6}{-0.0334619} & \cellcolor{gray!6}{-0.2470344} & \cellcolor{gray!6}{-0.0466343} & \cellcolor{gray!6}{Cloudy skies} & \cellcolor{gray!6}{Summer} & \cellcolor{gray!6}{19.30}\\
\cmidrule{1-9}
-0.0427 & 7.98e-01 & 0.0023737 & -0.0332559 & -0.2581092 & -0.0487201 & Cloudy skies & Autumn & 20.20\\
\bottomrule
\end{tabu}
\end{table}

@@

We have demonstrated that in the period 1993 - 2023, there is a positive
trend in SDR of (\(0.38\,\%/y\)) (brightening) (positive trend) in
Thessaloniki, Greece, under all-sky conditions. A previous study
\citep{Bais2013} for the period 1993 - 2011 found also a positive trend
of \(0.33\,\%/y\). The increase of this trend indicates that the
brightening of SDR continues and is probably caused by continuing
decreases in aerosol optical depth and the optical thickness of clouds
over the area. Moreover, we found a similar trend under clear-sky
conditions (\(0.16\,\%/y\)) that further supports the assumption that
the brightening is caused mainly by decreasing aerosols. Unfortunately,
for the entire period there is no available data for the aerosols, in
order to quantify their effect on SDR. However, \citet{Siomos2020} have
shown that aerosol optical depth over Thessaloniki is decreasing
constantly at least up to 2018. The attenuation of SDR by aerosols over
Europe have been proposed as major factor by \citet{Wild2021}. The
dimming effect on SDR under cloudy-sky conditions (\(0.019\,\%/y\)),
suggests that cloud optical thickness is decreasing during this period.
Because we have no adequate data to investigate the long term changes of
cloud thickness in the region, we cannot verify if the negative SDR
trend we observe under under cloudy-skies can be attributed solely to
changes in clouds.

The observed brightening on SDR over Thessaloniki is dependent on SZA
(larger SZAs lead to stronger brightening). The trend is also dependent
on season, with winter showing the strongest statistically significant
trend of \(0.69\) and \(0.7\,\%/y\) for all- and clear-skies,
respectively, in contrast to spring and summer. The trends for autumn
are also significant but smaller ( \(0.43\) and \(0.13\,\%/y\) for all-
and clear-skies, respectively). Our findings are in agreement with other
studies for the region.

>>>>

Due to the significant measurement uncertainty when the Sun is near the
horizon, we have excluded all measurements with SZA greater than
\(85^\circ\). Moreover, due to some obstructions around the site (hills
and buildings), we excluded data with azimuth angle between \(58^\circ\)
and \(120^\circ\) with SZA greater than \(78^\circ\). On the latter
instances, Sun is systematically not visible from the instrument's
location. To make the measurements comparable throughout the dataset, we
adjusted all one-minute radiometric values to the mean Sun - Earth
distance. Subsequently, we made all measurements relative to the Total
Solar Irradiance (TSI) at \(1\,\text{au}\), in order to compensate for
the Sun's intensity variability, using a time series of satellite TSI
observations. The TSI data we used are part of the ``NOAA Climate Data
Record of Total Solar Irradiance'' dataset \citep{Coddington2005}. The
initial daily values of this dataset were interpolated to match the time
step of our measurements. In the final dataset \(NA\%\) of the 1-minute
data were identified as under clear-sky conditions and \(NA\%\) as under
cloud-sky conditions.

@@

We calculated the linear SDR trends, from the departures of the mean
daily values from the daily climatology and for the three sky conditions
(Table~\ref{tab:trendtable}). In Figure~\ref{fig:trendALL} we present
only the time series under all-sky conditions; the plots for clear-sky
and cloud-sky conditions, are very similar and are shown in the Appendix
(Figures~\ref{fig:trendCLEAR} and~ \ref{fig:trendCLOUD}). We observe a
positive trend for all-sky conditions (\(0.38\,\%/y\)), a very close but
smaller trend for clear-skies (\(0.041\,\%/y\)) and a negative weaker
trend for cloudy-skies (\(0.34\,\%/y\)). In the studied period, there is
no significant break or change in the variability of the time series.
Other studies for the European region reported a change of the SDR
slope, around 1980 \citep{Wild2021, Yuan2021, Ohmura2009}, a few years
before the start of our records. It is interesting to note, that for the
observations period, the trend of the TSI is \(-0.00022\,\%/y\), and
thus we can not attribute any major effect on SDR trend to Solar
variability.

@@

\caption{\label{tab:trendtable}Trends in SDR daily means for different sky conditions for the period 1993 - 2023.}
\begin{tabu} to \linewidth {>{\centering\arraybackslash}p{8em}>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\centering}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\centering}X}
\toprule
slope & slope.p & Rsqrd & RsqrdAdj & cor.t & cor.estimate & DATA & N & slope.stat\_sig\\
\midrule
0.3760 & 0.0000 & 0.008339 & 0.0082421 & 9.284 & 0.09132 & All skies & 10251 & 100.00\\
0.3430 & 0.0000 & 0.005312 & 0.0051756 & 6.234 & 0.07289 & Cloudy skies & 7279 & 100.00\\
0.0406 & 0.0141 & 0.001195 & 0.0009971 & 2.455 & 0.03458 & Clear skies & 5036 & 98.59\\
\bottomrule
\end{tabu}
\end{table}

@@

Similarly to the long term trends discussed above, we have calculated
the trend of the anomalies for the three different sky conditions, and
for each season of the year, using the corresponding mean monthly values
(Figure~\ref{fig:seasonalALL} and Table~\ref{tab:trendseasontable}). For
all-sky conditions the trend in SDR in winter is the largest
(\(0.69\,\%/y\)), followed by the trend in autumn (\(0.43\,\%/y\), a
value close to the long term trend) both statistically significant above
the \(99\,\%\) confidence level. In spring and summer, the trends are
much smaller and of lesser statistical significance. These seasonal
differences indicate a possible relation of the trends in SDR to trends
of clouds during winter and autumn. For clear-skies, the trend in winter
is \(0.5\,\%/y\), larger than for all-skies (\(0.69\,\%/y\)), which is
another indication of a decreasing trend in cloud optical thickness.
Moreover, the trends under clear- and cloudy-sky conditions are almost
complementary to each other, particularly for winter and autumn, where
the signal is stronger. During spring and summer the statistical
significance is very low and the actual trend too small for a meaningful
comparison.

@@

\caption{\label{tab:trendseasontable}SDR linear trends of monthly anomalies for each season of the year.}
\begin{tabu} to \linewidth {>{\centering\arraybackslash}p{8em}>{\centering}X>{\raggedleft}X>{\raggedleft}X>{\centering}X>{\centering}X>{\raggedleft}X>{\raggedleft}X>{\centering}X}
\toprule
Trend [\%/year] & slope.p & Rsqrd & RsqrdAdj & cor.t & cor.estimate & Sky condition & Season & slope.stat\_sig\\
\midrule
\cellcolor{gray!6}{0.6870} & \cellcolor{gray!6}{0.001330} & \cellcolor{gray!6}{0.3121839} & \cellcolor{gray!6}{0.2876190} & \cellcolor{gray!6}{3.564906} & \cellcolor{gray!6}{0.5587342} & \cellcolor{gray!6}{All skies} & \cellcolor{gray!6}{Winter} & \cellcolor{gray!6}{99.9}\\
\cmidrule{1-9}
0.1460 & 0.178000 & 0.0616942 & 0.0293388 & 1.380858 & 0.2483832 & All skies & Spring & 82.2\\
\cmidrule{1-9}
\cellcolor{gray!6}{0.1230} & \cellcolor{gray!6}{0.096100} & \cellcolor{gray!6}{0.0957525} & \cellcolor{gray!6}{0.0634580} & \cellcolor{gray!6}{1.721910} & \cellcolor{gray!6}{0.3094390} & \cellcolor{gray!6}{All skies} & \cellcolor{gray!6}{Summer} & \cellcolor{gray!6}{90.4}\\
\cmidrule{1-9}
0.4300 & 0.006670 & 0.2347063 & 0.2073744 & 2.930403 & 0.4844650 & All skies & Autumn & 99.3\\
\cmidrule{1-9}
\cellcolor{gray!6}{0.4990} & \cellcolor{gray!6}{0.000166} & \cellcolor{gray!6}{0.4025975} & \cellcolor{gray!6}{0.3812617} & \cellcolor{gray!6}{4.343912} & \cellcolor{gray!6}{0.6345057} & \cellcolor{gray!6}{Clear skies} & \cellcolor{gray!6}{Winter} & \cellcolor{gray!6}{100.0}\\
\cmidrule{1-9}
-0.0865 & 0.315000 & 0.0348099 & 0.0015275 & -1.022690 & -0.1865741 & Clear skies & Spring & 68.5\\
\cmidrule{1-9}
\cellcolor{gray!6}{-0.2180} & \cellcolor{gray!6}{0.000398} & \cellcolor{gray!6}{0.3659586} & \cellcolor{gray!6}{0.3433142} & \cellcolor{gray!6}{-4.020094} & \cellcolor{gray!6}{-0.6049451} & \cellcolor{gray!6}{Clear skies} & \cellcolor{gray!6}{Summer} & \cellcolor{gray!6}{100.0}\\
\cmidrule{1-9}
0.1090 & 0.221000 & 0.0529350 & 0.0191112 & 1.251009 & 0.2300761 & Clear skies & Autumn & 77.9\\
\cmidrule{1-9}
\cellcolor{gray!6}{0.5200} & \cellcolor{gray!6}{0.009580} & \cellcolor{gray!6}{0.2164332} & \cellcolor{gray!6}{0.1884487} & \cellcolor{gray!6}{2.781012} & \cellcolor{gray!6}{0.4652238} & \cellcolor{gray!6}{Cloudy skies} & \cellcolor{gray!6}{Winter} & \cellcolor{gray!6}{99.0}\\
\cmidrule{1-9}
0.1780 & 0.122000 & 0.0806332 & 0.0489309 & 1.594819 & 0.2839598 & Cloudy skies & Spring & 87.8\\
\cmidrule{1-9}
\cellcolor{gray!6}{0.1940} & \cellcolor{gray!6}{0.036800} & \cellcolor{gray!6}{0.1466137} & \cellcolor{gray!6}{0.1161356} & \cellcolor{gray!6}{2.193277} & \cellcolor{gray!6}{0.3829017} & \cellcolor{gray!6}{Cloudy skies} & \cellcolor{gray!6}{Summer} & \cellcolor{gray!6}{96.3}\\
\cmidrule{1-9}
0.3470 & 0.100000 & 0.0934930 & 0.0611177 & 1.699350 & 0.3057662 & Cloudy skies & Autumn & 90.0\\
\bottomrule
\end{tabu}
\end{table}

@@

We have demonstrated that in the period 1993 - 2023, there is a positive
trend in SDR of (\(0.38\,\%/y\)) (brightening) (positive trend) in
Thessaloniki, Greece, under all-sky conditions. A previous study
\citep{Bais2013} for the period 1993 - 2011 found also a positive trend
of \(0.33\,\%/y\). The increase of this trend indicates that the
brightening of SDR continues and is probably caused by continuing
decreases in aerosol optical depth and the optical thickness of clouds
over the area. Moreover, we found a similar trend under clear-sky
conditions (\(0.041\,\%/y\)) that further supports the assumption that
the brightening is caused mainly by decreasing aerosols. Unfortunately,
for the entire period there is no available data for the aerosols, in
order to quantify their effect on SDR. However, \citet{Siomos2020} have
shown that aerosol optical depth over Thessaloniki is decreasing
constantly at least up to 2018. The attenuation of SDR by aerosols over
Europe have been proposed as major factor by \citet{Wild2021}. The
dimming effect on SDR under cloudy-sky conditions (\(0.34\,\%/y\)),
suggests that cloud optical thickness is decreasing during this period.
Because we have no adequate data to investigate the long term changes of
cloud thickness in the region, we cannot verify if the negative SDR
trend we observe under under cloudy-skies can be attributed solely to
changes in clouds.

The observed brightening on SDR over Thessaloniki is dependent on SZA
(larger SZAs lead to stronger brightening). The trend is also dependent
on season, with winter showing the strongest statistically significant
trend of \(0.69\) and \(0.5\,\%/y\) for all- and clear-skies,
respectively, in contrast to spring and summer. The trends for autumn
are also significant but smaller ( \(0.43\) and \(0.11\,\%/y\) for all-
and clear-skies, respectively). Our findings are in agreement with other
studies for the region.
